### 2012-06-12 trunk/ijcis-pos-model/sources/temporaldatabases.tex
preliminary final modifications

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Time
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\label{subsec:time-in-databases}Time in Databases}
The concept of time has been studied in databases for a long time. A true standard for adding temporal aspects to relational databases does not exist, but there is a consensus in the literature \cite{Dyreson1994} on what is called a \emph{temporal database}: a temporal database is a database dealing with some aspects of time in its schema.
In a temporal DBMS, a \textbf{chronon} is the shortest duration of time supported by the system. In temporal databases, some temporal attributes can be managed without treating the attribute differently from non-temporal attributes. The time described by such an attribute is called \textbf{user defined time} (\emph{UDT}). In addition to UDT, the following types of time can be discerned in a temporal database, all of which are handled exceptionally by the DBMS:

@@

	E.g., consider a database containing employee contract descriptions. The time when the employee's contract is valid, represented by an interval, is VT. The time when the employee's contract is stored in the database is the TT. The time when the decision for hiring this employee was made is the DT.

@@

%	\begin{itemize}	
%	\item \textbf{Granularity} denotes a partition on the set of chronons. The conversion between several granularities is studied in \cite{Lin97efficientconversion}. Granularity is the base of the temporal model in \cite{Cru97}. An object-oriented implementation is in \cite{624013}.
%	\item	To ensure \textbf{consistency}, temporal databases usually redefine the primary key of a relation. The new primary key takes into account the presence of the time. In order to keep the consistency, the DML is re-defined. For example, in a VTDB, the temporal update sentence is usually composed by an update sentence (closing the old row) plus an insert sentence (creating the new row).
%	\end{itemize}
%Guy's suggestion: instead of imprecision, use imperfection which is more general.
\subsubsection{Imperfection and time}
Representing imprecision and its semantics when dealing with time has been studied for a long time. Several proposals for representing and computing imprecise time indications can be found in \cite{DeCaluwe1997,DeTre1997}. Also, the changes between several granularities can be seen as a source of imprecision \cite{Devos1998}.

In the proposal section we will consider two kinds of imprecision:
\begin{itemize}
\item \textbf{Imperfection in the database} denotes the uncertainty that arises when the knowledge about the temporal data in the database is uncertain. E.g., a database record shows that \emph{`The car is in the garage around April.'}
 \item \textbf{Imprecision in the query specification} denotes the imprecision in the specification of temporal criteria by the user, when querying. E.g., \emph{`The user wants to obtain a car which is red and which is in the garage around April.'}
\end{itemize}

\subsubsection{Representation}
Several proposals for managing uncertain time in a database exist. Some proposals work with rough sets \cite{Qiang2009}, other proposals rely on possibility distributions for representing uncertainty in time \cite{Dyreson1998,Garrido2009,Galindo2001}. In order to compare temporal possibility distributions, extensions of the classical Allen's operators \cite{Allen1983} are defined in \cite{Ohlbach2004,Nagypal2003,Dubois2003a,Schockaert2008}.
%In the proposal section, we will follow the representation by means of possibility distributions, in order to work with both satisfaction and dissatisfaction degrees. Also, in order to work properly with fuzzy operators, the underlying domain should be numeric. 
% Take a look into this:
%In this paper, the representation for the dates will follow the Julian Day Number (JDN) representation \cite{Dir96}.

@@

In order to deal with uncertainty in time intervals, several proposals are made. Here, two approaches are introduced: the first one, based on \emph{Fuzzy Validity Period}~\cite{Garrido2009} and the second one based on \emph{Possibilistic Valid-time Period}~\cite{JoseEnriquePons2012}.

\begin{definition}
A \textbf{Fuzzy Validity Period} (\emph{FVP}) is defined as a fuzzy time interval specifying when the data regarding an object is valid. A fuzzy time interval is then the fuzzification of a crisp time interval.
\end{definition}
Several options to transform possibility distributions corresponding to the fuzzy starting point and the fuzzy end point into one consistent FVP exist \cite{Garrido2009}, e.g (Fig. \ref{fig:fuzzy-validity-period}):
\begin{itemize}
\item The \textbf{convex hull} approach is the most intuitive approach. The resulting FVP is the convex hull of the union of both fuzzy sets.
\item The \textbf{uncertainty preserving} approach is less intuitive but more realistic. The amount of uncertainty is maintained at the edges of the possibility distribution representing the FVP \cite{Garrido2009}.
\end{itemize}

@@

\begin{definition}
A \textbf{Possibilistic Valid-time Period} (\emph{PVP}) is an ill-known interval in time specifying when the data regarding an object is valid.
\end{definition}
A PVP is an ill-known interval in the sense of definition \ref{def;possibilistic-variable}, section \ref{subsec:possibilistic-variables}. Note that this representation is \emph{disjunctive}: the PVP represents only one crisp time interval, but that for some reason it is (partially) unknown.

The ill-known interval approach has many advantages from the representation by the FVP as demonstrated in \cite{Pons2011},\cite{Pons2012}. Table \ref{tbl:comparative-pvp-fvp} is a comparative between PVP and FVP. The following list defines the items in the comparative:

\begin{enumerate}
\item Domain: The domain of the possibility distribution modelled by the approach.
\item Implementation of relationships: How to implement a relationship.
\item Allen's relations: Are the Allen's relations defined?
\item Storage: The way the data is stored in the database.
\item Possibility measures: Does the framework provides always a possibility measure for any relation between the temporal elements?
\item Necessity measures:  Does the framework provides always a necessity measure for any relation between the temporal elements?
\end{enumerate}

@@

\subsection{\label{subsubsec:Understanding-valid-time-databases}Understanding Valid-time Databases}
This subsection is devoted to describe the behaviour of a crisp valid-time database. For the shake of simplicity, only the three main operations in the Data Manipulation Language \emph{DML} a.k.a \emph{CRUD} (CReate Update, Delete) operations are shown. Usually the DML operations in a temporal database are re-defined e.g. a typical update sentence in SQL could be re-defined by means of several insert and update sentences. Therefore, in order to be clear, these high level primitives in the DML for a valid-time database are usually noted as Insert, Modify and Delete. In the following subsections each primitive is defined and explained. Finally a illustrative example is given. For a more complete information on the behaviour of a bitemporal database, please refer to ~\cite{Jensen1994}.

\begin{definition}
Consider a relation $R$ in a temporal database, an entity $A$ given by the attributes $\left(a_1, \ldots, a_n \right)$ and the crisp time interval $I$ given by $\left[a,\ b\right] : a,b \in \T, a \leq b$. The pair $\left(A, I\right)$ expresses that the data regarding the entity A is valid during the crisp time interval I.  Note that the relation $R = \left(A, I \right) x \left(A, I \right)$. The following equation indicates that the pair $\left(A, I\right)$ is in the relation R:
\begin{equation}
\label{eq:rel-def}
\left( R, \left(A , I\right) \right)
\end{equation}
\end{definition}
\begin{example}
 Consider a library database. Each physical book is stored with an identifier (ID), a title and the start and end loan dates respectively. Table \ref{tbl:library-sample} illustrates the first version of the database, after three insertions. In this example, $A = ($ ID, Title $)$ and $I = ($ Start Loan, End Loan $)$.
\end{example}

\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:library-sample}Sample library database}
\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c c}\\
\hline
\textbf{ID}  & Title  & Start Loan & End Loan \\
\hline
3 & Dracula & 15/3/2012 & 30/3/2012 \\
4 & Frankenstein & 25/3/2012 & 4/4/2012 \\
5 & Harry Potter & 18/3/2012 & 2/4/2012\\
3 & Dracula & 4/4/2012 & UC \\
\hline\\
\end{tabular}
}

@@

\begin{definition}
 \label{def:until-changed}
\textbf{Until changed (UC):} To show that the data related to the entity $\left(A, I \right)$ is up-to-date, a special constant called \emph{Until Changed} is defined for the endpoint of the time interval $I$. The value for this constant is considered to be $+\infty$. 
\end{definition}

%quizas poner un ejemplo relacionado con lo de la liberia.

@@

For example, last row in Table \ref{tbl:library-sample}. The value for the time interval is $I = \left(4/4/2012 , UC \right)$. The meaning is that the book with ID = 3 was loaned the  4/4/2012 and it is still loaned. The book with ID = 3 is current in the relation.

\begin{definition}
\textbf{Current $\left(R, A \right)$:}
\label{def:current-in-relation}
This function returns a crisp time interval $I$ if the entity $A$ is current in the relation $R$ or $\emptyset$ if the entity $A$ is not current in the relation.

@@

\begin{align}
\label{eq:current-in-relation}
\mbox{Current} \left(R, A \right) &=& \\ 
\begin{cases}
\nonumber
I & \mbox{ if } (A,I) \in R \wedge I = \left(c, UC \right), c \in \T\\
\emptyset & \mbox{ in any other case }
\end{cases}
\end{align}
\end{definition}

For example, the function Current(R, (3,`Dracula')) returns the time interval $I = \left(4/4/2012 , UC  \right)$. On the other hand, Current(R, (4,`Frankenstein')) returns the empty set.

The following two Allen's relations (see Figure \ref{fig:allen-rel}) are defined like in~\cite{Nagypal2003} to implement the auxiliary functions.

@@

}
\end{center}
%\centerline{ \psfig{file=./graphs/Y-time-point.eps}}
\vspace*{10pt}
\fcaption{\label{fig:allen-rel}Allen's relations for two crisp time intervals $I$ and $J$.}
\vspace*{13pt}

\begin{definition}
\textbf{Overlaps:}
 \label{def:overlaps}
Given two time intervals $I = \left(c, d \right)$ and $J = \left(e, f\right)$, it is said that I overlaps J if:
\begin{equation}
 \label{eqn:overlaps}
\text{I Overlaps J}  = \left(c > e  \right) \wedge \left(d < f  \right) 
\end{equation}
\end{definition}

\begin{definition}
\textbf{During:}
 \label{def:during}
Given two time intervals $I = \left(c, d \right)$ and $J = \left(e, f\right)$, it is said that I during J if:
\begin{equation}
 \label{eqn:during}
\text{I During J}  = \left(c < e  \right) \wedge \left(d > e  \right) \wedge \left(d < f \right)
\end{equation}
\end{definition}

\begin{definition}
\textbf{CloseR$\left( I, J \right)$:}
\label{def:close-a-crisp-interval-r}
Consider two crisp intervals $I= \left[c ,d \right]$ and $J= \left[e ,f \right]$. The CloseR$\left(I, J\right)$ function allows to close the right-open interval $I$ with respect to the first value $e$ in $J$:
\begin{align}
\label{eq:close-a-crisp-interval}
\mbox{CloseR} \left( I, J \right) &=& \\ 
\begin{cases}
\nonumber
%I & \mbox{ if } b \neq +\infty \\
%I=\left[a, c \right[ & \mbox{ if } b = +\infty \wedge J > I
\left[c, e \right[ & \mbox{ if } d = UC \wedge J \mbox{ During } I \\
I & \mbox{ in any other case.}
\end{cases}
\end{align}
\end{definition}

@@

Now it is possible to close the current version of an entity by using \eqref{eq:close-a-crisp-interval} and \eqref{eq:current-in-relation}. This functionality is required to append new information about an existing entity in the relation.

\begin{definition}
\label{def:close-current-version}
Consider an entity $A$, the relation $R$ and a time interval $J$. Then, the function close-current$\left(R,\left(A,J\right) \right)$ closes any current version of the object $A$ if it exists and add the new version $\left(A, J \right)$.

\begin{eqnarray}
\label{eq:close-current}
\text{Close-current} \left(R, \left(A, J\right) \right) =\\
\begin{cases}
\nonumber
R - \left(A, I \right) \cup \left \lbrace \left(A, \mbox{ CloseR } \left(I, J\right) \right)\right \rbrace \cup \left \lbrace\left(A, J\right) \right \rbrace  \\
\nonumber
\mbox{ if } I = \mbox{ Current } \left(R, A \right)\\ %\wedge J > \mbox{ CloseR } \left(I, J \right)   \\
\nonumber R , \text{ in any other case}
\end{cases}
\end{eqnarray}
\end{definition}

For example, consider the book with ID = 3 in Table \ref{tbl:library-sample}. The current version was loaned on 4/4/2012. The book was returned to the library but, for some reason, the end loan date was not registered. The book is loaned again on 24/4/2012. Then, the Close-current function allows to close the current version of the book and insert a new version. First, the function CloseR is applied with $I = \left(4/4/2012, UC \right)$ and $J = \left(24/4/2012, UC \right)$. Hence, the value for the time interval $I$ is $ \left(4/4/2012, 23/4/2012 \right)$. Then, the modifications on the value for the time interval $I$ are stored. Finally, a new row with the current values of the book and the time interval $J$ is stored.
The result of this operation is illustrated on Table \ref{tbl:library-sample-2}.

@@

\begin{align}
\label{eq:modify}
\MOD \left( R, \left(A, I \right)\right) =\\
%\begin{cases}
\nonumber
\mbox{ Close-current }\left(R, A, \mbox{Current}\left(R, A \right) \right) %\\
% \mbox{ if Current}\left(R ,A \right) \neq \emptyset \\
% R,  \mbox{ otherwise }
%\end{cases}
\end{align}

\subsubsection{\label{subsubsec:insert}Insert}
The user wants to store an object $A$ which is valid in the relation $R$ during the time interval specified by $I = \left(a, b \right)$.
%
%
%The interpretation of the insert operation is the following: The user wants to store in the database that an object $A$  is true for some period(s) of time $I$ in the given relation $R$. To indicate that the object $A$ is current in the relation, the special value Until Current, $\UC$  is used. 
There are two main cases when performing a create operation:
\begin{enumerate}
\item The object $A$ was never in the relation $R$: The object is added with the valid-time indicated by the crisp interval $I$. In the example, this correspond with the following insertion sentences:
\begin{verbatim}
Insert(3,'Dracula',
(15/3/2012, 30/3/2012));
Insert(4,'Frankenstein',
(25/3/2012, 4/4/2012));
Insert(5,'Harry Potter',
(18/3/2012, 2/4/2012));
\end{verbatim}

\item The object $A$ is in the relation $R$. Depending on the value of the time interval $I$, there are three possibilities:
	\begin{enumerate}
	\item Insert $\left(A, I\right)$ in the relation $R$. If the time interval $I$ does not overlap any other valid-time interval in the relation $R$. For example, consider that the book with ID 3 is loaned on 4/4/2012 but the book has not been returned yet. This is illustrated in Table \ref{tbl:library-sample}. The insert sentence is:
	      \begin{verbatim}
Insert(3,'Dracula',
(4/4/2012, UC));
	      \end{verbatim}

	\item Modify and close the current version of $A$ and insert $\left(A, I\right)$. For example, consider now that the book with ID 3 is loaned  the 24/4/2012. Here the problem is that the book with ID 3 was loaned on 4/4/2012, but not returned. If the book is already in the library for a new loan, then it is necessary to set the return date and add a new row with the new loan date. This is illustrated in Table \ref{tbl:library-sample-2}, the insert sentence is:
	    \begin{verbatim}
Insert(3,'Dracula',
(24/4/2012, UC));
	      \end{verbatim}
	\item Reject the insertion, if the time interval $I$ does overlap any existing valid-time interval for the object $A$ in the relation $R$. For example, consider now that the librarian wants to introduce a past loan for the book with ID 3. The book was loaned on 6/4/12/2012 and returned on 25/4/2012. As this interval does overlaps other time intervals, it is not possible that the book was loaned in that time interval.Therefore, the insertion is rejected. The insert sentence is:

	      \begin{verbatim}
Insert(3,'Dracula',
(6/4/12/2012,25/4/2012));
	    \end{verbatim}

@@

\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:library-sample-2}Sample library database}
%\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c c}\\
\hline
\textbf{ID}  & Title  & Start Loan & End Loan \\
\hline
3 & Dracula & 15/3/2012 & 30/3/2012 \\
4 & Frankenstein & 25/3/2012 & 4/4/2012 \\
5 & Harry Potter & 18/3/2012 & 2/4/2012\\
3 & Dracula & 4/4/2012 & 23/4/2012 \\
3 & Dracula & 24/4/2012 & UC \\
\hline\\
\end{tabular}
}

@@

\begin{align}
\label{eq:insert}
\INS \left(R ,\left(A, I\right)\right) =\\
\begin{cases}
\nonumber
R \cup \left(A, I \right) & \mbox{ if }  (A, I ) \not \in R  \vee \\
 &   \left(A, J\right) \in R \wedge \lnot \left(I \mbox{ Overlaps } J\right) \\
\mbox{modify(R,(A,I))} & \mbox{ otherwise }    %\mbox{ if } \exists J \left(A,J\right) \in R \wedge \mbox{Current(A,J)} \\ 
%R, &   
\end{cases} 	
\end{align}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Time
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\label{subsec:time-in-databases}Time in Databases}
The concept of time has been studied in the database framework for a long time. A true standard for adding temporal aspects to relational databases does not exist, but there is a consensus in the literature \cite{Dyreson1994} on what is called a \emph{temporal database}: a temporal database is a database dealing with some aspects of time in its schema.
In a temporal DBMS, a \textbf{chronon} is the shortest duration of time supported by the system. In temporal databases, some temporal attributes can be managed without treating the attribute differently from non-temporal attributes. The time described by such an attribute is called \textbf{user defined time} (\emph{UDT}). In addition to UDT, the following types of time can be discerned in a temporal database, all of which are handled exceptionally by the DBMS:

@@

	For example, consider a database containing employee contract descriptions. The time when the employee's contract is valid, represented by an interval, is the VT. The time when the employee's contract is stored in the database is the TT. The time when the decision for hiring this employee was made is the DT.

@@

%	\begin{itemize}	
%	\item \textbf{Granularity} denotes a partition on the set of chronons. The conversion between several granularities is studied in \cite{Lin97efficientconversion}. Granularity is the base of the temporal model in \cite{Cru97}. An object-oriented implementation is in \cite{624013}.
%	\item	To ensure \textbf{consistency}, temporal databases usually redefine the primary key of a relation. The new primary key takes into account the presence of the time. In order to keep the consistency, the DML is re-defined. For example, in a VTDB, the temporal update sentence is usually composed by an update sentence (closing the old row) plus an insert sentence (creating the new row).
%	\end{itemize}
%Guy's suggestion: instead of imprecision, use imperfection which is more general.
\subsubsection{Imperfection and time}
The representation of imprecision and its semantics when dealing with time has been studied for a long time. Several proposals for representing and handling imprecise time indications can be found in \cite{DeCaluwe1997,DeTre1997}. Also, the changes between several granularities can be seen as a source of imprecision \cite{Devos1998}.

In this paper we will consider two kinds of imperfection:
\begin{itemize}
\item \textbf{Imperfection in the database} the knowledge about the temporal data contains some imperfection. E.g., a database record shows that \emph{`The car was in the garage around April.'}
 \item \textbf{Imprecision in the query specification} denotes the imprecision in the specification of temporal criteria by the user, when querying. E.g., \emph{`The user wants a car which was in the garage around April.'}
\end{itemize}

\subsubsection{Representation}
Several proposals for managing uncertain time in a database exist. Some proposals work with rough sets \cite{Qiang2009}, and some others rely on possibility distributions for representing uncertainty in time \cite{Dyreson1998,Garrido2009,Galindo2001}. In order to compare temporal possibility distributions, the extensions of the classical Allen's operators \cite{Allen1983} are defined in \cite{Ohlbach2004,Nagypal2003,Dubois2003a,Schockaert2008}.
%In the proposal section, we will follow the representation by means of possibility distributions, in order to work with both satisfaction and dissatisfaction degrees. Also, in order to work properly with fuzzy operators, the underlying domain should be numeric. 
% Take a look into this:
%In this paper, the representation for the dates will follow the Julian Day Number (JDN) representation \cite{Dir96}.

@@

In order to deal with uncertainty in time intervals, several proposals have been made. Here, two approaches are introduced: the first one, based on \emph{Fuzzy Validity Periods}~\cite{Garrido2009} and the second one, based on \emph{Possibilistic Valid-time Periods}~\cite{JoseEnriquePons2012}.

\begin{definition}
A \textbf{Fuzzy Validity Period} (\emph{FVP}) is defined as a fuzzy time interval specifying when the data regarding an object are valid. A fuzzy time interval is then the fuzzification of a crisp time interval.\\
\end{definition}

Several options to transform possibility distributions corresponding to the fuzzy starting point and the fuzzy end point into one consistent FVP exist \cite{Garrido2009}, e.g (Fig. \ref{fig:fuzzy-validity-period}):
\begin{itemize}
\item The \textbf{convex hull} approach is the most intuitive approach. The resulting FVP is the convex hull of the union of both possibility distributions.
\item The \textbf{uncertainty preserving} approach. The amount of uncertainty is maintained at the edges of the possibility distribution representing the FVP \cite{Garrido2009}.
\end{itemize}

@@

The main feature for the FVP is the optimization for the storage. The compact representation is the result of a conjunctive semantic. The object is valid within all the time points inside the starting and ending points.

\begin{definition}
A \textbf{Possibilistic Valid-time Period} (\emph{PVP}) is an ill-known interval of time specifying when the data regarding an object are valid.
\end{definition}
A PVP is an ill-known interval in the sense of definition \ref{def;possibilistic-variable}. Note that this representation is \emph{disjunctive}: the PVP represents only one crisp time interval, but for some reason, it is (partially) unknown.

The main feature for the PVP is that it preserves all the information for both starting and ending points\cite{Pons2011,Pons2012}. Table \ref{tbl:comparative-pvp-fvp} is a comparative between PVP and FVP. The following list defines the items in the comparative:

\begin{enumerate}
\item Domain: The domain of the possibility distribution modelled by the approach.
\item Implementation of relationships: How to implement a relationship.
\item Allen's relations: Are the Allen's relations defined?
\item Storage: The way the data are stored in the database.
\item Possibility measures: Does the framework provide always a possibility measure for any relation between the temporal elements?
\item Necessity measures:  Does the framework provide always a necessity measure for any relation between the temporal elements?
\end{enumerate}

@@

\subsection{\label{subsubsec:Understanding-valid-time-databases}Understanding Valid-time Databases}
This subsection is devoted to describe the behaviour of a crisp valid-time database. For the sake of simplicity, only the three main operations in the Data Manipulation Language (CReate Update, Delete) are shown. Usually the DML operations in a temporal database are re-defined (a typical update sentence in SQL could be re-defined by means of several insert and update sentences). Therefore, for the sake of clarity, these high level primitives in the DML for a valid-time database are usually noted as Insert, Modify and Delete. In the following subsections each primitive is defined and explained. Finally, an illustrative example is given. For a more complete information on the behaviour of a bitemporal database, please refer to ~\cite{Jensen1994}.

\begin{definition}
\emph{Valid-time relation.}
Consider $A=\left(A_1, \ldots, A_n \right)$ be the set of attributes that define an entity and $I = \left(s_i,e_i \right)$ the set of attribute defining the time interval.  The pair $\left(A, I\right)$ expresses that the data regarding the entities A are valid during the crisp time interval I. Let $R \subseteq A\  x\  I$ be a valid-time relation. Then, the following equation indicates that the pair $\left(A, I\right)$ is in the relation R:

\begin{equation}
\label{eq:rel-def}
\left( R, \left(A , I\right) \right)
\end{equation}
\end{definition}

@@

\begin{definition}
 \label{def:from-beginning}
\emph{From the beginning (FB).} 
Consider $A=\left(A_1, \ldots, A_n \right)$ be the set of attributes that define an entity and $I = \left(s_i,e_i \right)$ the set of attribute defining the time interval. Let $R \subseteq A\  x\  I$ be a valid-time relation. The from the beginning constant is used when the value for $s_i$ is equal to $-\infty$.
\end{definition}

\begin{definition}
 \label{def:until-changed}
\emph{Until changed (UC).} 
Consider $A=\left(A_1, \ldots, A_n \right)$ be the set of attributes that define an entity and $I = \left(s_i,e_i \right)$ the set of attribute defining the time interval. Let $R \subseteq A\  x\  I$ be a valid-time relation. The until changed constant is used when the value for $e_i$ is equal to $+\infty$.
\end{definition}

@@

For example, last row in Table \ref{tbl:library-sample}. The value for the time interval is $I = \left(4/4/2012 , UC \right)$. The meaning is that the patient with ID = 3 was in the hospital the  4/4/2012 and it is still there. The patient with ID = 3 is current in the relation.

\begin{example}
\label{ex:library-database}
 Consider a hospital database. Each patient is stored in the database with an identifier (ID), the name and the dates when the patient entered and left the hospital. Table \ref{tbl:library-sample} illustrates the first version of the database, after three insertions. In this example, $A = ($ ID, Name $)$ and $I = ($ Arrival, Leave $)$.
\end{example}

\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:library-sample}Sample hospital database}
%\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c c}\\
\hline
\textbf{ID}  & Name  & Arrival & Leave \\
\hline
3 & John & 15/3/2012 & 30/3/2012 \\
4 & Antoon & 25/3/2012 & 4/4/2012 \\
5 & Daan & 18/3/2012 & 2/4/2012\\
3 & John & 4/4/2012 & UC \\
\hline\\
\end{tabular}
}

\begin{definition}
\label{def:current-in-relation}
\emph{Current $\left(R, A \right)$.}
Consider $A=\left(A_1, \ldots, A_n \right)$ be the set of attributes that define an entity and $I = \left(s_i,UC \right)$ the set of attribute defining the time interval. Let $R \subseteq A\  x\  I$ be a valid-time relation. The function Current $\left(R, A \right)$ returns the crisp time interval $I$ if the entity $A$ is currently in the relation $R$ or the empty set, if the entity $A$ is not currently in the relation.

@@

\begin{align}
\label{eq:current-in-relation}
\mbox{Current} \left(R, A \right) &=& \\ 
\begin{cases}
\nonumber
I & \mbox{ if } (A,I) \in R \mbox{ and } I = \left(c, UC \right), c \in \T\\
\emptyset & \mbox{ in any other case }
\end{cases}
\end{align}
\end{definition}

For example, the function Current(R, (3,`John')) returns the time interval $I = \left(4/4/2012 , UC  \right)$. On the other hand, Current(R, (4,`Antoon')) returns the empty set.

The Allen's relations between two time intervals are shown in Figure \ref{fig:allen-rel}. The complete implementation of the relations using only the starting and the ending points is defined in~\cite{Nagypal2003}.

@@

}
\end{center}
%\centerline{ \psfig{file=./graphs/Y-time-point.eps}}
\vspace*{10pt}
\fcaption{\label{fig:allen-rel}Allen's relations for two crisp time intervals $I$ and $J$. Note that the crisp time interval $I$ is fixed and the different positions of the interval $J$ illustrate the Allen's relations.}
\vspace*{13pt}

\begin{definition}
\emph{Overlaps:}
 \label{def:overlaps}
Given two time intervals $I = \left(c, d \right)$ and $J = \left(e, f\right)$, it is said that I overlaps J if:
\begin{equation}
 \label{eqn:overlaps}
\text{I Overlaps J}  = \left(c > e  \right) \wedge \left(d < f  \right) 
\end{equation}
\end{definition}

\begin{definition}
\emph{During:}
 \label{def:during}
Given two time intervals $I = \left(c, d \right)$ and $J = \left(e, f\right)$, it is said that I during J if:
\begin{equation}
 \label{eqn:during}
\text{I During J}  = \left(c < e  \right) \wedge \left(d > e  \right) \wedge \left(d < f \right)
\end{equation}
\end{definition}

In the rest of the paper, and without losing generality, we will consider that the time granularity are days. Also, the dates will be given in the format \emph{dd/mm/yyyy}.

\begin{definition}
\emph{CloseR$\left( I, J \right)$:}
\label{def:close-a-crisp-interval-r}
Consider two crisp intervals $I= \left[c ,d \right]$ and $J= \left[e ,f \right]$. The CloseR$\left(I, J\right)$ function allows to close the right-open interval $I$ with respect to the first value $e$ in $J$:
\begin{align}
\label{eq:close-a-crisp-interval}
\mbox{CloseR} \left( I, J \right) &=& \\ 
\begin{cases}
\nonumber
%I & \mbox{ if } b \neq +\infty \\
%I=\left[a, c \right[ & \mbox{ if } b = +\infty \wedge J > I
\left[c, e \right[ & \mbox{ if } d = UC \mbox{ and } J \mbox{ During } I \\
I & \mbox{ in any other case.}
\end{cases}
\end{align}
\end{definition}

@@

Now it is possible to close the current version of an entity by using \eqref{eq:close-a-crisp-interval} and \eqref{eq:current-in-relation}. This functionality is required to add or update new information about an existing entity in the relation.

\begin{definition}
\label{def:close-current-version}
\emph{Close-current$\left(R, \left(A, J  \right)\right)$}.
Consider $A=\left(A_1, \ldots, A_n \right)$ be the set of attributes that define an entity and $J = \left(s_j,e_j \right)$ the set of attribute defining the time interval. Let $R \subseteq A\  x\  I$ be a valid-time relation. Then, the function close-current$\left(R,\left(A,J\right) \right)$ closes any current version of the object $A$ if it exists and add the new version $\left(A, J\right)$.

\begin{eqnarray}
\label{eq:close-current}
\text{Close-current} \left(R, \left(A, J\right) \right) =\\
\begin{cases}
\nonumber
R - \left(A, I \right) \cup \left \lbrace \left(A, \mbox{ CloseR } \left(I, J\right) \right)\right \rbrace \cup \left \lbrace\left(A, J\right) \right \rbrace  \\
\nonumber
\mbox{ if  Current } \left(R, A \right) \neq \emptyset\\ %\wedge J > \mbox{ CloseR } \left(I, J \right)   \\
\nonumber R , \text{ in any other case}
\end{cases}
\end{eqnarray}
\end{definition}

For example, consider the patient with ID = 3 in Table \ref{tbl:library-sample}. The current version arrived on 4/4/2012. The patient left the hospital but, for some reason, the date  was not registered. The patient arrived again at the hospital on 24/4/2012. Then, the Close-current function allows to close the current version of the patient and insert a new version. First, the function CloseR is applied with $I = \left(4/4/2012, UC \right)$ and $J = \left(24/4/2012, UC \right)$. Hence, the value for the time interval $I$ is $ \left(4/4/2012, 23/4/2012 \right)$. Then, the modifications on the value for the time interval $I$ are stored. Finally, a new row with the current values of the patient and the time interval $J$ is stored.
The result of this operation is illustrated on Table \ref{tbl:library-sample-2}.

@@

\begin{align}
\label{eq:modify}
\MOD \left( R, \left(A, I \right)\right) =\\
%\begin{cases}
\nonumber
\mbox{ Close-current }\left(R, \left(A, I\right) \right) %\\
% \mbox{ if Current}\left(R ,A \right) \neq \emptyset \\
% R,  \mbox{ otherwise }
%\end{cases}
\end{align}

\subsubsection{\label{subsubsec:insert}Insert}
The user wants to store an object $A$ which is valid in the relation $R$ during the time interval specified by $I = \left(a, b \right)$.
%
%
%The interpretation of the insert operation is the following: The user wants to store in the database that an object $A$  is true for some period(s) of time $I$ in the given relation $R$. To indicate that the object $A$ is current in the relation, the special value Until Current, $\UC$  is used. 
There are two main cases when performing a create operation:
\begin{enumerate}
\item The object $A$ was never in the relation $R$: The object is added with the valid-time indicated by the crisp interval $I$. 
% In the example, this correspond with the following insertion sentences:
% \begin{verbatim}
% Insert(3,'Dracula',
% (15/3/2012, 30/3/2012));
% Insert(4,'Frankenstein',
% (25/3/2012, 4/4/2012));
% Insert(5,'Harry Potter',
% (18/3/2012, 2/4/2012));
% \end{verbatim}

\item The object $A$ is in the relation $R$. Depending on the value of the time interval $I$, there are three possibilities:
	\begin{enumerate}
	\item Insert $\left(A, I\right)$ in the relation $R$. If the time interval $I$ does not overlap any other valid-time interval in the relation $R$. For example, consider that the book with ID 3 is loaned on 4/4/2012 but the book has not been returned yet. This is illustrated in Table \ref{tbl:library-sample}.
%  The insert sentence is:
% 	      \begin{verbatim}
% Insert(3,'Dracula',
% (4/4/2012, UC));
% 	      \end{verbatim}

	\item Modify and close the current version of $A$ and insert $\left(A, I\right)$. For example, consider now that the book with ID 3 is loaned  the 24/4/2012. Here the problem is that the book with ID 3 was loaned on 4/4/2012, but not returned. If the book is already in the library for a new loan, then it is necessary to set the return date and add a new row with the new loan date. This is illustrated in Table \ref{tbl:library-sample-2}.
%  the insert sentence is:
% 	    \begin{verbatim}
% Insert(3,'Dracula',
% (24/4/2012, UC));
% 	      \end{verbatim}
	\item Reject the insertion, if the time interval $I$ does overlap any existing valid-time interval for the object $A$ in the relation $R$. For example, consider now that the librarian wants to introduce a past loan for the book with ID 3. The book was loaned on 6/4/12/2012 and returned on 25/4/2012. As this interval does overlaps other time intervals, it is not possible that the book was loaned in that time interval.Therefore, the insertion is rejected.
%  The insert sentence is:
% 
% 	      \begin{verbatim}
% Insert(3,'Dracula',
% (6/4/12/2012,25/4/2012));
% 	    \end{verbatim}

@@

\vglue13pt
%\begin{table}[htbp]
\tcap{\label{tbl:library-sample-2}Sample library database}
%\centerline{\small DATA TYPES}
\vglue-6pt
\centerline{\small\baselineskip=13pt
\begin{tabular}{c c c c}\\
\hline
\textbf{ID}  & Name  & Arrival & Leave \\
\hline
3 & John & 15/3/2012 & 30/3/2012 \\
4 & Antoon & 25/3/2012 & 4/4/2012 \\
5 & Daan Potter & 18/3/2012 & 2/4/2012\\
3 & John & 4/4/2012 & 23/4/2012 \\
3 & John & 24/4/2012 & UC \\
\hline\\
\end{tabular}
}

@@

\begin{align}
\label{eq:insert}
\INS \left(R ,\left(A, I\right)\right) =\\
\begin{cases}
\nonumber
R \cup \left(A, I \right) & \mbox{ if }  (A, I ) \not \in R  \mbox{ or } \\
 &   \left(A, J\right) \in R \mbox{ and } \lnot \left(I \mbox{ Overlaps } J\right) \\
\mbox{modify(R,(A,I))} & \mbox{ otherwise }    %\mbox{ if } \exists J \left(A,J\right) \in R \wedge \mbox{Current(A,J)} \\ 
%R, &   
\end{cases} 	
\end{align}

\subsubsection{\label{subsubsec:del}Delete}
The delete operation logically removes a current entity $A$ which is valid in the relation $R$:
\begin{align}
\label{eq:delete}
\DEL \left(R ,A \right) =\\
\begin{cases}
\nonumber
R - \left \lbrace \left(A, I_1 \right),  \ldots, \left(A, I_1 \right)   \right \rbrace, & \forall I_i : \left( A, I_i \right) \in R\\
R, & \mbox{ otherwise }  
\end{cases} 	
\end{align}
